---
title: "Fenceline Map Data Analysis and Visualization"
author: "Paige Varner"
date: "2024-02-09"
output: html_document
---

#load libraries
```{r}
library("dplyr")
library("bigrquery")
library("DBI")
library("pheatmap")
library("RColorBrewer")
library("Rmisc")
library("stats")
library("rstatix")
library("ggplot2")
library("tigris")
library("sf")
library("tidycensus")
library("viridis")
```

#read and format data
```{r}
data = read.csv("~/TSCA/Fenceline Map/fencelinemap/mapdata_v4.csv")

#add weighted risk levels
data$cancer_risklevel = cut(data$Cancer_weighted_5yr_sum, c(-Inf,20000,100000,Inf), c("lower", "higher", "highest"))

data$asthma_risklevel = cut(data$Asthma_weighted_5yr_sum, c(-Inf,20000,100000,Inf), c("lower", "higher", "highest"))

data$dev_risklevel = cut(data$Dev_weighted_5yr_sum, c(-Inf,20000,100000,Inf), c("lower", "higher", "highest"))

#subset into health endpoints
cancer = filter(data, Cancer_PoundsReleased_5yr_sum > 0)
asthma = filter(data, Asthma_PoundsReleased_5yr_sum > 0)
dev = filter(data, Dev_PoundsReleased_5yr_sum > 0)
```


#percent of total facilities releasing more than 1 chemical that causes cancer, asthma, or dev
```{r}
#cancer: 336/1683 = 20%
nrow(cancer[grepl(", ", cancer$Chemical), ])

#dev: 338/1115 = 30%
nrow(dev[grepl(", ", dev$Chemical), ])

#asthma: 150/890  = 17%
nrow(asthma[grepl(", ", asthma$Chemical), ])

```


#number of people exposed to 1 or more of cancer, asthma, dev causing chemicals from same facility (within 6 miles)
```{r}
#cancer: 39661270
cancer2 = filter(cancer[grepl(", ", cancer$Chemical), ])
sum(cancer2$population_10km)

#dev: 36499048
dev2 = filter(dev[grepl(", ", dev$Chemical), ])
sum(dev2$population_10km)

#asthma: 14075491
asthma2 = filter(asthma[grepl(", ", asthma$Chemical), ])
sum(asthma2$population_10km)
```
#percent of total facilities neighboring another facility (within 6 miles)
```{r}
#all facilities: 1237/2116 = 58%
sum(data$F10km_facnum > 0) 

#cancer: 965/1683 = 57%
sum(cancer$F10km_facnum > 0) 

#asthma: 445/890 = 50%
sum(asthma$F10km_facnum > 0) 

#dev: 743/1115 = 67%
sum(dev$F10km_facnum > 0)
```


#number of people living with in 6 miles of more than 1 facility
```{r}
#all facilities: 179303362
data6 = subset(data, data$F10km_facnum > 0)
sum(data6$population_10km)

#cancer: 134336644
cancer6 = subset(cancer, cancer$F10km_facnum > 0)
sum(cancer6$population_10km)

#asthma: 50917757
asthma6 = subset(asthma, asthma$F10km_facnum > 0)
sum(asthma6$population_10km)

#dev: 111839393
dev6 = subset(dev, dev$F10km_facnum > 0)
sum(dev6$population_10km)

```


#percent of total facilities that contribute to mroe than 1 health effect 
```{r}
# 1464/2116 = 69%
sum(data$Health_Risk_Count > 1)
```

#total population exposed to facilities contributing to more than 1 health effect
```{r}
# 129243466
health2 = subset(data, data$Health_Risk_Count > 1)
sum(health2$population_10km)
```


#percent of total facilities that are either releasing more than 1 chemical contributing to cancer, asthma, or dev, OR is neighboring at least 1 more facility
```{r}
#all facilities: 1341/2116 = 63%
sum(grepl(", ", data$Chemical) | data$F10km_facnum > 0)

#cancer: 1066/1683 = 63%
sum(grepl(", ", cancer$Chemical) | cancer$F10km_facnum > 0) 

#asthma: 488/890 = 55%
sum(grepl(", ", asthma$Chemical) | asthma$F10km_facnum > 0)

#dev: 833/1115 = 75%
sum(grepl(", ", dev$Chemical) | dev$F10km_facnum > 0)
```


#population with 6 miles of facilities that are either releasing more than 1 chemical contributing to cancer, asthma, or dev, OR is neighboring at least 1 more facility
```{r}
#all facilities: 183859376
data3 = subset(data, grepl(", ", data$Chemical) | data$F10km_facnum > 0)
sum(data3$population_10km)

#cancer: 138420463
cancer3 = subset(cancer, grepl(", ", cancer$Chemical) | cancer$F10km_facnum > 0) 
sum(cancer3$population_10km)

#asthma: 52663270
asthma3 = subset(asthma, grepl(", ", asthma$Chemical) | asthma$F10km_facnum > 0)
sum(asthma3$population_10km)

#dev: 115897240
dev3 = subset(dev, grepl(", ", dev$Chemical) | dev$F10km_facnum > 0)
sum(dev3$population_10km)
```


## adding baseline vulnerability data to CEAM data ##

#get census tract GEOIDs for all facilities based on long/lat
```{r}
cvi = read.csv("~/TSCA/CRA/CRA_Prioritization/Data/cvi_scores.csv")
colnames(cvi)[1] = "GEOID"

#getting census tract info for all TRI facilities
tracts = tracts(state = NULL, cb = TRUE)

data.points <- data %>%
  st_as_sf(coords=c('Longitude', 'Latitude'), crs=st_crs(tracts))

data.tracts = st_join(data.points, tracts)
```


#join CVI data to CEAM data based on GEOID (why am I losing a bunch of rows here?)
```{r}
data_cvi = merge(data.tracts, cvi, by = "GEOID")
```

#put categorical CVI values into another column
```{r}
data_cvi$bv_cat = cut(data_cvi$Baseline.Vulnerabilities, c(-Inf, 0.344, 0.489, 0.643, 0.751, Inf), c("very low", "low", "medium", "high", "very high"))

```

#subset data_cvi by health risk
```{r}
cancer_cvi = filter(data_cvi, Cancer_PoundsReleased_5yr_sum > 0)
asthma_cvi = filter(data_cvi, Asthma_PoundsReleased_5yr_sum > 0)
dev_cvi = filter(data_cvi, Dev_PoundsReleased_5yr_sum > 0)
```


#correlations between risk levels and vulnerability levels overall: not seeing much of a correlation
```{r}
#baseline vul ~ cancer risk
reg_cvi_vul = lm(data_cvi$Baseline.Vulnerabilities ~ data_cvi$Cancer_weighted_5yr_sum)
reg_cancer = ggplot(data_cvi, aes(x = log(Cancer_weighted_5yr_sum), y = Baseline.Vulnerabilities)) + geom_point() + geom_smooth(method = "lm", se = FALSE)
reg_cancer

#baseline vul ~ asthma risk
reg_asthma = ggplot(data_cvi, aes(x = log(Asthma_weighted_5yr_sum), y = Baseline.Vulnerabilities)) + geom_point() + geom_smooth(method = "lm", se = FALSE)
reg_asthma

#baseline vul ~ dev risk
reg_dev = ggplot(data_cvi, aes(x = log(Dev_weighted_5yr_sum), y = Baseline.Vulnerabilities)) + geom_point() + geom_smooth(method = "lm", se = FALSE)
reg_dev
```


#summary of risk levels for each vulnerability, graph, (could still run statistics if I want/need)
```{r}
#cancer
bvlevel_summary_cancer = summarySE(cancer_cvi, groupvars = "bv_cat", measurevar = "Cancer_weighted_5yr_sum")

bvlevel_cancer = ggplot(bvlevel_summary_cancer, aes(x = bv_cat, y = Cancer_weighted_5yr_sum)) + geom_point(size = 3) + theme_bw() + ylab("Weighted Sum of 5 Year Release (lbs)") + xlab("Baseline Vulnerability Level") + theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) + geom_errorbar(aes(ymin = Cancer_weighted_5yr_sum - se, ymax = Cancer_weighted_5yr_sum + se), width = 0.2)

bvlevel_cancer

ggsave(filename = "bv_risklevel_cancer.png", device = "png", plot = bvlevel_cancer, path = "~/TSCA/Fenceline Map/fencelinemap/Output", width = 10, height = 5)

#cancer stats
cancer.aov = aov(Cancer_weighted_5yr_sum ~ bv_cat, data = cancer_cvi)
cancer.aov
cancer.tukey = TukeyHSD(x = cancer.aov, conf.level = 0.95)
cancer.tukey


#asthma
bvlevel_summary_asthma = summarySE(asthma_cvi, groupvars = "bv_cat", measurevar = "Asthma_weighted_5yr_sum")

bvlevel_asthma = ggplot(bvlevel_summary_asthma, aes(x = bv_cat, y = Asthma_weighted_5yr_sum)) + geom_point(size = 3) + theme_bw() + ylab("Weighted Sum of 5 Year Release (lbs)") + xlab("Baseline Vulnerability Level") + theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) + geom_errorbar(aes(ymin = Asthma_weighted_5yr_sum - se, ymax = Asthma_weighted_5yr_sum + se), width = 0.2)

bvlevel_asthma

ggsave(filename = "bv_risklevel_asthma.png", device = "png", plot = bvlevel_asthma, path = "~/TSCA/Fenceline Map/fencelinemap/Output", width = 10, height = 5)

#dev
bvlevel_summary_dev = summarySE(dev_cvi, groupvars = "bv_cat", measurevar = "Dev_weighted_5yr_sum")

bvlevel_dev = ggplot(bvlevel_summary_dev, aes(x = bv_cat, y = Dev_weighted_5yr_sum)) + geom_point(size = 3) + theme_bw() + ylab("Weighted Sum of 5 Year Release (lbs)") + xlab("Baseline Vulnerability Level") + theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) + geom_errorbar(aes(ymin = Dev_weighted_5yr_sum - se, ymax = Dev_weighted_5yr_sum + se), width = 0.2)

bvlevel_dev

ggsave(filename = "bv_risklevel_dev.png", device = "png", plot = bvlevel_dev, path = "~/TSCA/Fenceline Map/fencelinemap/Output", width = 10, height = 5)
```


#correlating high vulnerability and high risk with heat map (none of these are interesting and are counter to what we would expect)
```{r}
#cancer
cat_combos_cancer = prop.table(table(cancer_cvi$cancer_risklevel, cancer_cvi$bv_cat)) * 100
print(cat_combos_cancer)

cancer_cats_matrix = as.matrix(cat_combos_cancer)

pheatmap(cat_combos_cancer, cluster_rows = FALSE, cluster_cols = FALSE, display_numbers = FALSE, cellheight=32,cellwidth=32, fontsize_number = 10, angle_col = 45, legend = T, filename = "~/TSCA/Fenceline Map/fencelinemap/Output/catheatmap_cancer.png", width = 5, height = 3)

#asthma
cat_combos_asthma = prop.table(table(asthma_cvi$asthma_risklevel, asthma_cvi$bv_cat)) * 100
print(cat_combos_asthma)
asthma_cats_matrix = as.matrix(cat_combos_asthma)

pheatmap(cat_combos_asthma, cluster_rows = FALSE, cluster_cols = FALSE, display_numbers = FALSE, cellheight=32,cellwidth=32, fontsize_number = 10, angle_col = 45, legend = T, filename = "~/TSCA/Fenceline Map/fencelinemap/Output/catheatmap_asthma.png", width = 5, height = 3)

#dev
cat_combos_dev = prop.table(table(dev_cvi$dev_risklevel, dev_cvi$bv_cat)) * 100
print(cat_combos_dev)
dev_cats_matrix = as.matrix(cat_combos_dev)

pheatmap(cat_combos_dev, cluster_rows = FALSE, cluster_cols = FALSE, display_numbers = FALSE, cellheight=32,cellwidth=32, fontsize_number = 10, angle_col = 45, legend = T, filename = "~/TSCA/Fenceline Map/fencelinemap/Output/catheatmap_dev.png", width = 5, height = 3)
```


#top 10 cities/counties with the highest exposures 
```{r}
#all facilities
data_cities = summarySE(data, groupvars = c("City", "State"), measurevar = "PoundsReleased_5yr_sum", na.rm = TRUE)
data_top10 = data_cities %>%
  arrange(desc(PoundsReleased_5yr_sum))

#cancer
cancer_cities = summarySE(cancer, groupvars = c("City", "State"), measurevar = "PoundsReleased_5yr_sum", na.rm = TRUE)
cancer_top10 = cancer_cities %>%
  arrange(desc(PoundsReleased_5yr_sum))

#asthma
asthma_cities = summarySE(asthma, groupvars = c("City", "State"), measurevar = "PoundsReleased_5yr_sum", na.rm = TRUE)
asthma_top10 = asthma_cities %>%
  arrange(desc(PoundsReleased_5yr_sum))

#dev
dev_cities = summarySE(dev, groupvars = c("City", "State"), measurevar = "PoundsReleased_5yr_sum", na.rm = TRUE)
dev_top10 = dev_cities %>%
  arrange(desc(PoundsReleased_5yr_sum))
```


#top 10 cities/counties with the highest weighted risk levels
```{r}
#cancer
cancer_cities_risk = summarySE(cancer, groupvars = c("City", "State"), measurevar = "Cancer_weighted_5yr_sum", na.rm = TRUE)
cancer_top10_risk = cancer_cities_risk %>%
  arrange(desc(Cancer_weighted_5yr_sum))

write.csv(cancer_top10_risk, "~/TSCA/Fenceline Map/fencelinemap/Output/cancer_topcities_risk.csv")


#asthma
asthma_cities_risk = summarySE(asthma, groupvars = c("City", "State"), measurevar = "Asthma_weighted_5yr_sum", na.rm = TRUE)
asthma_top10_risk = asthma_cities_risk %>%
  arrange(desc(Asthma_weighted_5yr_sum))

write.csv(asthma_top10_risk, "~/TSCA/Fenceline Map/fencelinemap/Output/asthma_topcities_risk.csv")

#dev
dev_cities_risk = summarySE(dev, groupvars = c("City", "State"), measurevar = "Dev_weighted_5yr_sum", na.rm = TRUE)
dev_top10_risk = dev_cities_risk %>%
  arrange(desc(Dev_weighted_5yr_sum))

write.csv(dev_top10_risk, "~/TSCA/Fenceline Map/fencelinemap/Output/dev_topcities_risk.csv")
```


##Black Americans impact analysis##

#across the US: average % of Black and white in 10 km and in county; numbers of individuals
```{r}
#add new columns of the percent change from 10 km to county for white and Black (making it so that if the value is positive, there's a higher percentage of the population near facilities, and if negative, there's a lower percentage near facilities/10 km)

data$Blk_difference = 100 * ((data$BlkPercent_10km - data$BlkPercent_county)/data$BlkPercent_county)

data$Wht_difference = 100 * ((data$WhtPercent_10km - data$WhtPercent_county)/data$WhtPercent_county)

data$Blk_difference_dec = data$Blk_difference / 100
data$Wht_difference_dec = data$Wht_difference / 100

#ttest: p < 2.2e-16; mean of Blk difference is 17%, mean of Wht difference is -4.6%
black.white.ttest = t.test(data$Blk_difference, data$Wht_difference, paired = FALSE)
black.white.ttest

#for odds ratio calculation
sum(data$Blk_pop_10km) #36031778
sum(data$Blk_pop_county) #261624205
sum(data$WhtPercent_10km * data$population_10km) #129611327
sum(data$WhtPercent_county * data$populationE_county) #1056687894

#adding columns to get number of Black americans at 10 km and county; percent of Black population that lives near facilities
data$Blk_pop_10km = data$population_10km * data$BlkPercent_10km
data$Blk_pop_county = data$populationE_county * data$BlkPercent_county

data$Blk_10km_over_county = (data$Blk_pop_10km / data$Blk_pop_county)


#number of Black americans living near these facilities = 36031778
sum(data$Blk_pop_10km)


```


#filtering for location-specific statistics
```{r}
#Houston area
houston = data %>%
  filter(State == "TX") %>%
  filter(County == "HARRIS" | County == "MONTGOMERY" | County == "LIBERTY" | County == "CHAMBERS" | County == "GALVESTON" | County == "BRAZORIA" | County == "FORT BEND" | County == "WALLER")

#cancer alley
cancer.alley = data %>%
  filter(State == "LA") %>%
  filter(County == "IBERVILLE PARISH" | County == "WEST BATON ROUGE PARISH" | County == "EAST BATON ROUGE PARISH" | County == "ASCENSION PARISH" | County == "ST JAMES PARISH" | County == "ST. JOHN THE BAPTIST PARISH" | County == "LAFAYETTE PARISH" | County == "JEFFERSON PARISH" | County == "ORLEANS PARISH")

#detroit
detroit = data %>%
  filter(State == "MI") %>%
  filter(County == "WAYNE" | County == "MACOMB")

#chicago
chicago = data %>%
  filter(State == "IL" | State == "IN") %>%
  filter(County == "COOK" | County == "DUPAGE" | County == "LAKE")

#LA
la = data %>%
  filter(State == "CA") %>%
  filter(County == "LOS ANGELES" | County == "ORANGE")

```


#difference at these locations between white/Black near facilities (10km) compared to whole county
```{r}
#houston: p < 3.698e-5; mean of Blk difference is -20.3%%, mean of Wht difference is 5%
houston.ttest = t.test(houston$Blk_difference, houston$Wht_difference, paired = FALSE)
houston.ttest

#cancer alley: p < 4.951e-8; mean of Blk difference is 31%, mean of Wht difference is -18.8%
cancer.alley.ttest = t.test(cancer.alley$Blk_difference, cancer.alley$Wht_difference, paired = FALSE)
cancer.alley.ttest

#cancer alley odds ratio numbers, then done online
sum(cancer.alley$Blk_pop_10km) #1218749
sum(cancer.alley$Blk_pop_county) #2593216
sum(cancer.alley$WhtPercent_10km * cancer.alley$population_10km) #892658
sum(cancer.alley$WhtPercent_county * cancer.alley$populationE_county) #3413085

#detroit: p = 0.02432; mean of Blk difference is -21.5%, mean of Wht difference is 8.61%
detroit.ttest = t.test(detroit$Blk_difference, detroit$Wht_difference, paired = FALSE)
detroit.ttest

#chicago: no significant difference
chicago.ttest = t.test(chicago$Blk_difference, chicago$Wht_difference, paired = FALSE)
chicago.ttest

#la: p = no significant difference
la.ttest = t.test(la$Blk_difference, la$Wht_difference, paired = FALSE)
la.ttest
```


#percent of the Black population in these areas that live near facilities; number of Black people near facilities
```{r}
#houston = on average, 23.4% of Black Americans living in Houston area live within 6 miles of at least 1 facility; this is 2030340 individuals
mean(houston$Blk_10km_over_county)
sum(houston$Blk_pop_10km)

#cancer alley = on average, 78.5% of Black Americans living near cancer alley live within 6 miles of at least 1 facility; this is 1218749 individuals
mean(cancer.alley$Blk_10km_over_county)
sum(cancer.alley$Blk_pop_10km)

#detroit = on average, 11.1% of Black Americans living in detroit area live within 6 miles of at least 1 facility; this is 1448516 individuals
mean(detroit$Blk_10km_over_county)
sum(detroit$Blk_pop_10km)

#chicago = on average, 10.6% of Black Americans living in chicago area live within 6 miles of at least 1 facility; this is 3091814 individuals
mean(chicago$Blk_10km_over_county)
sum(chicago$Blk_pop_10km)

#LA = on average, 9.5% of Black Americans living in LA area live within 6 miles of at least 1 facility; this is 1646540 individuals
mean(la$Blk_10km_over_county)
sum(la$Blk_pop_10km)
```





